---
title: "TARDIS 1.1.2 run guidelines (August 2026)"
author: Eewon Tai <Eewon.Tai@UGent.be>
output: html_notebook
---

TARDIS v1.1.2 should be downloaded, installed, loaded, and ran in the following order. The basic idea is to 1) download package from Github repository 2) install & load it via devtools 3) run createTargetList and tardisPeaks in sequence. More details are given below.

## Download package from Github repository

1.  Go to the repository page: <https://github.com/UGent-LIMET/TARDIS/tree/main>
2.  On the top right: Click on Code -\> Local -\> Download ZIP
3.  Wait for the download to complete
4.  Go to that zip file location, unzip it in a specific directory of your choice (this is where your TARDIS code is)
5.  Now we have a folder of the TARDIS package.

## Install and load all necessary packages

First, we need to install all necessary packages. After installing each of them, we load them. Important: run it at the beginning every time you start TARDIS. This block has to be run first before all the other code in this script.

```{r}
if (!require("tidyr")) install.packages("tidyr")
library(tidyr)
if (!require("dplyr")) install.packages("dplyr")
library(dplyr)
if (!require("xcms")) install.packages("xcms")
library(xcms)
if (!require("Spectra")) install.packages("Spectra")
library(Spectra)
if (!require("devtools")) install.packages("devtools")
library(devtools)
```

Important: if you encounter error messages (not warnings) while doing this, you should not ignore them! Try to find out why that error occurred, fix the error, and re-run the script until the error disappears, which means it ran successfully. You need to have every package here installed and loaded successfully before proceeding to the next step.

## Install & load TARDIS package via devtools

Next, we install and load the TARDIS package with devtools. This is the preferred way due to its simplicity and speed.

1.  If you are installing TARDIS the first time in your computer: just install and load it.

```{r}
# install
devtools::install("C:/Users/TARDIS-v1.1.2/")  # example path to the TARDIS folder (created in the first step of this script)
# load
devtools::load_all("C:/Users/TARDIS-v1.1.2/")  # example path to the TARDIS folder (created in the first step of this script)
```

2.  If you have changed the code in the TARDIS package, or you want to use another version of the package instead of the already existing one: you need to remove the existing TARDIS package first and then install & load the next one, to avoid conflict. If you don't remove the existing TARDIS package, sometimes the code just runs the old code.

```{r}
remove.packages("TARDIS")
# install
devtools::install("C:/Users/TARDIS-v1.1.2/")  # example path to the TARDIS folder (created in the first step of this script)
# load
devtools::load_all("C:/Users/TARDIS-v1.1.2/")  # example path to the TARDIS folder (created in the first step of this script)
```

If you have installed the package once in your computer (devtools::install), it will not go away unless you explicitly delete it. It stays in the C drive library folder. Installing is only necessary once at the beginning. Then, you only need to load the package (devtools::load_all) to use it from next time.

Important: if you encounter error messages (not warnings) while doing this, you should not ignore them! Try to find out why that error occurred, fix the error, and re-run the script until the error disappears, which means it ran successfully.

## Run code

- Run createTargetList function and tardisPeaks function in the same way as TARDIS v1.1.0 by Dr. Pablo Vangeenderhuysen, described in more detail here: <https://pablovgd.github.io/TARDIS/>
